---
title: "HW5_TFang"
author: "Tianyi Fang"
date: "October 29, 2017"
output:
  word_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Gradient descent for blind source separation
####1.Plot p(y)
```{r}
set.seed(1)
y <- sort(rnorm(100, 0,1))
p <- function(y) {1/(pi*cosh(y))}
set.seed(2)
g <- dnorm(y, 0, sqrt(pi/2))
plot(y, p(y), type = "l", ylab = "p(y)/Guassian", main = "Plot of p(y) and Guassian")
points(y, g, type = "l", col = "red")
legend("topright", inset = 0.05, title = "Distributions",c("p(y)","Guassian"), lwd =2, lty = c(1,1),col = c("black", "red"))


```

####2.show 
$logp(X|W)=Tlog(|W|) + \sum_{t=1}^{T}\sum_{s=1}^{3}logp(w^sx_t)$.
$logP(X|W) = log(\prod_{t}^{T}P(x_t|A^{-1}))$

$=log(\prod_{t}^{T}\frac{1}{|A^{-1}|}P_y(\frac{x}{A^{-1}}))=$

$log(\prod_{t}^{T}{|W|}P_y({Wx_t}))=$


$log(\prod_{t}^{T} \prod_{s}^{3}|W|P_y(w^sx_t)) = \sum_{t}^{T}log|W| + \sum_{t}^{T}\sum_{s}^{3}logP_y(w^sx_t)$

$= Tlog|W|+ \sum_{t}^{T}\sum_{s}^{3}logP_y(w^sx_t)$

####3.Show that
$Wadj(W) = det(W)I$, $\frac{adj(w)}{det(W)} = \frac{I}{W} = W^{-1}$

$\frac{\partial log(|W|)}{\partial W_{ij}} =$ ,

$\frac{1}{det(W)}adj^{T}(W_{ij}) = W_{ij}^{-1} = a_{ji}$

$\frac{\partial \sum_{t=1}^{T}\sum_{s=1}^{3}P(w^sx_t)}{\partial w_{ij}} = \frac{\partial \sum_{t=1}^{T}P(w_{ij}x_{jt})}{\partial y_{it}}x_{jt}$ =
$\frac{\partial \sum_{t=1}^{T}P(Y_{it})}{\partial y_{it}}x_{jt}$
For $w_{ij}x_{jt} = y_{it}, w_{ij} = \frac{y_{it}}{x_{jt}}$


####4.Plug in the expression for the gradient to write the update rule.
$W_{new} = W_old + \eta A^T_{old} + \frac{1}{T}(-tanh(Y)X^T))$
####5. Function of gradient descent
```{r}
gradient_descent <- function (matrix, dl){
  #initialize w
  t <- dim(x)[2]#4900000
  set.seed(3)
  aa <- rnorm(9,0,1)
  A_old <- t(matrix(aa, ncol=3, nrow = 3))
  W_new <- solve(A_old) #W=A^{-1}
  #learning rate 
  eta <- 1
  #stop criteria:Flag
  #max likelihood estimate
  mle <-c()
  #iteration
  iteration = 0
  stop_value = 10
  while(stop_value > 0.000001){
    iteration = iteration + 1
    W_old <- W_new
    Y_old <- W_old %*% x
    W_new <- W_old + eta*(t(A_old) + (1/t) *(-tanh(Y_old)%*%t(x)))
    eta = eta*dl
    Y_new <- W_new %*% x
    A_old <- solve(W_new)
    mle[iteration] = t*log(abs(det(W_new))) + sum(log(1/(pi*cosh(Y_new))))
    stop_value = abs(sum(W_new - W_old))
  }
  W_hat <- W_new %*% solve(sqCov)
  return(list(W_new, mle, iteration, W_hat))
}
```
####loadin X
```{r}
library(audio)
library(dplyr)
library(ggplot2)
X <- matrix(0, ncol=490000, nrow = 3)
w1 <- load.wave("mike1.wav")
w2 <- load.wave("mike2.wav")
w3 <- load.wave("mike3.wav")
X <- rbind(w1,w2,w3)
```
####6. Write a 3x3 covariance matrix of X.
```{r}
cov <- matrix(0, ncol=3, nrow=3)
t <- dim(X)[2]
for(i in (1:3)){
  for(j in (1:3)){
    cov[i,j] <- 1/t*(t(X[i,])%*%X[j,])-(1/t^2)*(sum(X[i,]*sum(X[j,])))
  }
}
print(cov)
#plot

plot(X[1,],X[2,], xlab = "X1", ylab = "X2", main = "Scatterplot of X1,X2")
plot(X[2,],X[3,], xlab = "X2", ylab = "X3", main = "Scatterplot of X2,X3")
plot(X[1,],X[3,], xlab = "X1", ylab = "X3", main = "Scatterplot of X1,X3")
```
####7. get sqCov
```{r}
library(expm)
sqCov <- sqrtm(cov)
X_white <- solve(sqCov)%*%X
cov_white <- matrix(0, ncol=3, nrow=3)
t_white <- dim(X_white)[2]
for(i in (1:3)){
  for(j in (1:3)){
    cov_white[i,j] <- 1/t*(t(X_white[i,])%*%X_white[j,])-(1/t^2)*(sum(X_white[i,]*sum(X_white[j,])))
  }
}
cov_white
#almost =1
```
If we calculate the covariance matrix of X_white, we can see the covariance matrix of X_white is approximately identity matrix.

####8.Run gradient_descent on white data. How to initialize W? set eta, stop_sign, get the # of iteration, Plot the evoluation of log-likelihood. What?
```{r}
gradient_descent <- function (x, dl){
  #initialize w
  t <- dim(x)[2]#4900000
  set.seed(3)
  aa <- rnorm(9,0,1)
  A_old <- t(matrix(aa, ncol=3, nrow = 3))
  W_new <- solve(A_old) #W=A^{-1}
  #learning rate 
  eta <- 1
  #stop criteria:Flag
  #max likelihood estimate
  mle <-c()
  #iteration
  iteration = 0
  stop_value = 10
  while(stop_value > 0.000001){
    iteration = iteration + 1
    W_old <- W_new
    Y_old <- W_old %*% x
    W_new <- W_old + eta*(t(A_old) + (1/t) *(-tanh(Y_old)%*%t(x)))
    eta = eta*dl
    Y_new <- W_new %*% x
    A_old = solve(W_old)
    mle[iteration] = t*log(abs(det(W_new))) + sum(log(1/(pi*cosh(Y_new))))
    stop_value = abs(sum(W_new - W_old))
    A_old = solve(W_old)
    #print(W_new)
  }
  W_hat <- W_new %*% solve(sqCov)
  return(list(W_new, mle, iteration, W_hat))
}

#
white_0.9 <- gradient_descent(X_white, 0.9)
white_0.7 <- gradient_descent(X_white, 0.7)
white_0.5 <- gradient_descent(X_white, 0.5)
#W_new_0.9 = white_0.9[[1]]
mle_white_0.9 <-white_0.9[[2]]
iteration_0.9 <- white_0.9[[3]]
W_hat_white_0.9 <- white_0.9[[4]]
#W_new_0.7 = white_0.9[[1]]
mle_white_0.7 <-white_0.7[[2]]
iteration_0.7 <- white_0.7[[3]]
W_hat_white_0.7 <- white_0.7[[4]]
#W_new_0.5 = white_0.5[[1]]
mle_white_0.5 <-white_0.5[[2]]
iteration_0.5 <- white_0.5[[3]]
W_hat_white_0.5 <- white_0.5[[4]]
```
\item How initialize W
I initialized my W as the inverse of A, where A is a 3x3 matrix contain random normally distributed values from 0:1
\item How set eta/stop_cretiera
I set my learning rate as 1 and every iteration I decrease my learning rate by 10%, that is, every iteration it will descent in a smaller steps. I also tried 0.7, 0.5 to see whether how the steps of learning rate will affect the converge trend of mle.\\
My stop cretiera is: when W_new has little different from W_old, which means converge and get the min point, so we stop.
\item How many iteration
For eta=0.9\*eta, eta=0.7\*eta, eta=0.5\*eta 
```{r}
ite <- c(iteration_0.9, iteration_0.7, iteration_0.5)
ite
```
So we can see the larger step learning rate change, the less iteration time it will need to converge.
\item Plot evolution of mle
```{r}
mle <-as.data.frame(mle_white_0.9)
mle$iteration <- c(1:108)
mle$eta0.7 <- c(mle_white_0.7, rep_len(max(mle_white_0.7), 108-39))
mle$eta0.5 <- c(mle_white_0.5, rep_len(max(mle_white_0.5), 108-21))
colnames(mle)<- c("eta0.9","eta0.7","eta0.5")
library(reshape2)
#mle.melt <- melt(mle, id.var = "iteration")
#mle.melt%>%
 # ggplot(aes(x=iteration, y = value, color = variable)) + geom_line(size = 2) + labs(y="Max Likelihood Estimate",title = "Different Change rate of Learning rate vs MLE")
```
To sum up, with larger step of learning rate, The converge will be faster but the value of MLE is less than smaller step pf learning rate. For example, if learning rate reduce 0.5 every time, it is easily to go over the min MLE, and end up with local optima.
\item Return the W_hat
```{r}
print(W_hat_white_0.5)
print(W_hat_white_0.7)
print(W_hat_white_0.9)
```
####plot a histogram of raw, white data, recover data, Here I only plot the result of eta =eta\*0.9
```{r}
#raw data
X.df <- as.data.frame(t(X)) 
colnames(X.df) <- c("X1","X2","X3")
X.df %>% ggplot(aes(X1))+geom_histogram()+ labs(title = "Histogram of X1")
X.df %>% ggplot(aes(X2))+geom_histogram()+ labs(title = "Histogram of X2")
X.df %>% ggplot(aes(X3))+geom_histogram()+ labs(title = "Histogram of X3")
#White data
X_white.df <- as.data.frame(t(X)) 
colnames(X_white.df) <- c("X_white1","X_white2","X_white3")
X_white.df %>% ggplot(aes(X_white1))+geom_histogram()+ labs(title = "Histogram of X_white1")
X_white.df %>% ggplot(aes(X_white2))+geom_histogram()+ labs(title = "Histogram of X_white2")
X_white.df %>% ggplot(aes(X_white3))+geom_histogram()+ labs(title = "Histogram of X_white3")
#recovered data
recover_Y <- W_hat_white_0.9 %*% X
recover.df <- as.data.frame(t(recover_Y)) 
colnames(recover.df) <- c("Y_recover1","Y_recover2","Y_recover3")
recover.df %>% ggplot(aes(Y_recover1))+geom_histogram()+ labs(title = "Histogram of Y_recover1")
recover.df %>% ggplot(aes(Y_recover2))+geom_histogram()+ labs(title = "Histogram of Y_recover2")
recover.df %>% ggplot(aes(Y_recover3))+geom_histogram()+ labs(title = "Histogram of Y_recover3")
```
####10. Plot marginals for each source, What is the covarience?Plot three pairwise scatterplots
```{r}
library(ggExtra)
#marginal of Y1,Y2
p1 <- ggplot(recover.df, aes(recover.df$Y_recover1, recover.df$Y_recover2))+geom_point()+ labs(title = "Histogram of Marginal of Y_recover1 and Y_recover2")
ggMarginal(p1, type = "histogram") 

#marginal of Y2,3
p2 <- ggplot(recover.df, aes(recover.df$Y_recover2, recover.df$Y_recover3))+geom_point()+ labs(title = "Histogram of Marginal of Y_recover2 and Y_recover3")
ggMarginal(p1, type = "histogram") 

#marginal of Y1,3
p3 <- ggplot(recover.df, aes(recover.df$Y_recover1, recover.df$Y_recover3))+geom_point()+ labs(title = "Histogram of Marginal of Y_recover1 and Y_recover3")
ggMarginal(p1, type = "histogram") 
#cov
cov_y <- matrix(0, ncol=3, nrow=3)
t_y <- dim(recover_Y)[2]
for(i in (1:3)){
  for(j in (1:3)){
    cov_y[i,j] <- 1/t*(t(recover_Y[i,])%*%recover_Y[j,])-(1/t_y^2)*(sum(recover_Y[i,]*sum(recover_Y[j,])))
  }
}
cov_y
#
plot(recover_Y[1,],recover_Y[2,], xlab = "Y1", ylab = "Y2", main = "Scatterplot of Y1,Y2")
plot(recover_Y[2,],recover_Y[3,], xlab = "Y2", ylab = "Y3", main = "Scatterplot of Y2,Y3")
plot(recover_Y[1,],recover_Y[3,], xlab = "Y1", ylab = "Y3", main = "Scatterplot of Y1,Y3")
```
####11.Is the MLE unique? Explain 
```{r}
norm_y <- matrix(0, nrow= 3, ncol = dim(X)[2])
for(i in 1:3){
  norm_y[i,] <- recover_Y[i,]/(2*max(recover_Y[i,]))
}
save.wave(norm_y[1,], where = "C:\\Users\\Tianyi Fang\\Desktop\\stat545\\hw5\\recover1.wav")
save.wave(norm_y[2,], where = "C:\\Users\\Tianyi Fang\\Desktop\\stat545\\hw5\\recover2.wav")
save.wave(norm_y[3,], where = "C:\\Users\\Tianyi Fang\\Desktop\\stat545\\hw5\\recover3.wav")
```
MLE of W is not always unique. For a convex-optimization problem, it will always find the global optimum, which is unique. But in general, It also depends on the choices of your step size(learning rate), the initial set of W. If you want to find the global optimum, that is a NP -hard problem. Unless you tried as many as possible, you cannot decide whether it is local or global optima.   


