---
title: "STAT545_HW2"
author: "Tianyi Fang"
date: "9/17/2017"
output:
  html_document: default
  pdf_document:
    latex_engine: xelatex
---

## Problem 1
## Q1
Assuming the matrix A is a rank n matrix, it has n linearly independent eigenvectors, which form an eigenspace. The initial vector b can be written as a linear combination of the n eigenvectors, and when b (c_n being the eignspace coefficients) is multiplied to the matrix A, from the two equations, 
$$Ax = \lambda x$$
$$A(c_1 x_1 + c_2 x_2 + ... + c_n x_n) = \lambda_1 c_1 x_1 + \lambda_2 c_2 x_2 + ... + \lambda_n c_n x_n$$
we can see that whenever b is multiplied to A once and gets normalized, the coefficient of the dominant eigenvector (x_1) grows bigger and becomes closer to 1, while the others decrease to zero. This is why the power method with Rayleigh quotient gives the largest eigenvalue and eigenvector.  


## Q2
The condition number is a measurement of how much the output value of a function can change for a small change in the input. Associated with matrix, we have the Ax=b as the function, and we have the condition number as the change in solution x over the change in b (we denote as e here),
$$
\max_{e,b\ne0}\frac{\frac{\|A^{-1}e\|}{\|A^{-1}b\|}}{\frac{\|e\|}{\|b\|}} = 
\max_{e,b\ne0}(\frac{\|A^{-1}e\|}{\|e\|})(\frac{\|b\|}{\|A^{-1}b\|})=
\max_{e\ne0}(\frac{\|A^{-1}e\|}{\|e\|})\max_{x\ne0}(\frac{\|Ax\|}{\|x\|}) = \|A^{-1}\|\cdot\|A\|
$$
As we know, the matrix norm (spectral norm) of a matrix is its largest singular value, and since matrix A is symmetric, we know that the norm will be its largest eigenvalue. Also, the A-inverse has its eigenvalue as reciprocal as the eigenvalues of A (i.e. smallest eigenvalue of A has the largest reciprocal form). As a result, the condition number of a symmetric matrix is the ratio of its largest and smallest eigenvalues.  


## Problem 2
## Q3
Given a node with index i, its parent node is with index floor(i/2) or i%/%2, and its left child is with index 2xi, while right child is with index 2xi+1

## Q4
```{r}
# create a heap
make_heap <- function(LMAX){
  # Since it is a max heap, we use -inf here
  myheap = rep_len(-Inf, LMAX)
  heap_length = 0
  # return heap itself and the heap length
  return(list("heap" = myheap, "len" = heap_length))
}
```

## Q5
```{r}
# return max without deletion
find_max <- function(myheap,heap_length)
  if (heap_length==0) print("The heap is empty") else return(myheap[1])
```

## Q6
```{r}
delete_max <- function(myheap, heap_length){
  if (heap_length==0){
    print("The heap is empty")
    return()
  }
  # swap the max with the last element and discard the max (set to default)
  myheap[c(1,heap_length)] = myheap[c(heap_length,1)]
  mx = myheap[heap_length]
  myheap[heap_length] = -Inf
  heap_length = heap_length-1
  i = 1
  #print(myheap)
  # maintain the representation invariant (RI)
  # propagation downwards the tree
  while(2*i<=heap_length){
    if(2*i+1<=heap_length){
      if((myheap[2*i]>myheap[i]) | (myheap[2*i+1]>myheap[i])){
        if (myheap[2*i]>myheap[2*i+1]){
          myheap[c(i, 2*i)] = myheap[c(2*i, i)]
          i = 2*i
        }else{
          myheap[c(i, 2*i+1)]=myheap[c(2*i+1, i)]
          i = 2*i+1
        }
      }else break
    }else if(myheap[2*i]>myheap[i]){
      myheap[c(i, 2*i)] = myheap[c(2*i, i)]
      i = 2*i
    }else break
  }
  return(list("heap" = myheap, "len" = heap_length, "max" = mx))
}
```

## Q7
```{r}
insert <- function(myheap, hl, ii){
  # insert in the last and run RI for upward propagation
  # hl - heap_length, ii - inserted element
  myheap[hl+1] = ii
  #print(myheap)
  i = hl+1
  # propagration upward the tree
  while(i>1){
    if(myheap[i]>myheap[i%/%2]){
      myheap[c(i, i%/%2)] = myheap[c(i%/%2, i)]
      i = i%/%2
    }else break
  }
  return(list("heap" = myheap, "len" = hl+1))
}
```

## Q8
```{r}
heapsort <- function(n=20){
  # initilization of the heap
  hps = make_heap(n)
  hp = unlist(hps["heap"],use.name=FALSE)
  hpl = unlist(hps["len"],use.name=FALSE)
  # create n = 20 random number
  i = n
  # loop for sequential input
  while(i>0){
    # range can change, accordingly, as long as whithin (-inf, inf)
    ii = runif(1, -1000, 1000)
    hps = insert(hp, hpl, ii)
    hp = unlist(hps["heap"],use.name=FALSE)
    hpl = unlist(hps["len"],use.name=FALSE)
    i = i-1
  }
  # loop for sequential output max
  ret = rep_len(0, n)
  while(i<n){
    hps = delete_max(hp, hpl)
    hp = unlist(hps["heap"],use.name=FALSE)
    hpl = unlist(hps["len"],use.name=FALSE) 
    ret[i+1] = unlist(hps["max"], use.names=FALSE)
    i = i+1
  }
  return(ret)
}
heapsort(20)
```

## Problem 3
## Q1-2
```{r}
# Assume we have the same-length for w and v as the problem addressed
# All items are of positive weights and values & W_knapsack is integer
myKP <- function(w, v, W_knapsack){
  # boundary check
  if((length(w)==0) | (W_knapsack<=0)){
    return(list("val"=0, "obj"=c()))
  }
  # create a vector to store the optimal value for subproblems
  sub = rep_len(0, W_knapsack)
  sub_obj = rep_len(0, W_knapsack)
  # sort w to speed up the loop, need change v values accordingly
  ws = sort(w, index.return=TRUE)
  w = unlist(ws["x"], use.names=FALSE)
  w_idx = unlist(ws["ix"], use.names=FALSE)
  v = v[w_idx]
  #print(w)
  #print(v)
  # use topological sort order
  for(i in 1:W_knapsack){
    mymax = 0
    for(j in 1:length(w)){
      if (i-w[j]>=0){
        # tackle the base case (the first item (j) added or not)
        if (i-w[j]==0){
          if(mymax<v[j]){
            mymax = v[j]
            sub_obj[i] = j
            }
          # tackle the subproblem transition from sub[i-w[j]] -> sub[i] using max
        }else{
          if (mymax<v[j]+sub[i-w[j]]){
            mymax = v[j]+sub[i-w[j]]
            # record the jth item with its index
            sub_obj[i] = j
          }
        }
      # need not loop anymore since any weight comes after is larger
      }else{
        sub[i] = mymax
        break
      }
      sub[i] = mymax
    }  
  }
  #print(sub)
  #print(sub_obj)
  return(list("val"=sub, "obj"=sub_obj))
}
# test with examples
w = c(3,2,1,4,5,6)
v = c(10,8,1,10,19,25)
sol = myKP(w,v,25)
# There may be multiple solutions, only return one here
val = unlist(sol["val"],use.names=FALSE)
myobj = unlist(sol["obj"], use.names=FALSE)
cur = which(val==max(val))[1]
print(cur)
pack = rep_len(0, length(w))
while(cur>0){
  pack[myobj[cur]] = pack[myobj[cur]]+1
  cur = cur - w[myobj[cur]]
}
print("The optimal value is: ")
print(max(val))
print("The items that get picked: ")
print(pack)
```