---
title: "Stats 545 HW3"
author: "Tianyi Fang"
date: "September 28,2017"
output: pdf_document
---
##question 1
#1,
```{r setup, include=FALSE}
load_image_file = function(filename){
    ret = list()
    f = file(filename, 'rb') 
    readBin(f, 'integer', n=1,size=4,endian = 'big')
    ret$n = readBin(f,'integer',n=1,size=4,endian='big')
    nrow = readBin(f,'integer',n=1,size=4,endian='big')
    ncol = readBin(f,'integer',n=1,size=4,endian='big')
    x = readBin(f,'integer',n=ret$n*nrow*ncol,size=1,signed=F)
    ret$x = matrix(x, ncol=nrow*ncol, byrow=T)
    close(f)
    ret
  }
load_label_file = function(filename){
     f = file(filename,'rb')
     readBin(f,'integer',n=1,size=4,endian='big')
     n = readBin(f,'integer',n=1,size=4,endian='big')
     y = readBin(f,'integer',n=n,size=1,signed=F)
     close(f)
     y
}
train <<- load_image_file('C:/Users/Tianqiao/Desktop/study/purdue/17fall/STAT545 intro to computational stat/homework/train-images-idx3-ubyte/train-images.idx3-ubyte')
test <<- load_label_file('C:/Users/Tianqiao/Desktop/study/purdue/17fall/STAT545 intro to computational stat/homework/train-labels-idx1-ubyte/train-labels.idx1-ubyte')  


show_digit = function(arr784, col = gray(12:1/12), ...){
  image(matrix(arr784,nrow = 28)[,28:1], col = col, ...)
}
digits = train$x[1:1000,]
labels = test[1:1000]
show_digit(digits[5,])
```

#2, include (a), (b), (c)
```{r}
max = 255 #maximum value of one point one dim
#function computing distance between a vector and each row of a matrix
dist_vec2mat = function(x,mat){
  apply(mat,1,function(r) sum((r - x)^2))
}
#function computing every row of input_matrix
dist_mat2mat = function(input_mat, ref_mat) {
  dist_mat = apply(input_mat, 1, function(r) dist_vec2mat(r, ref_mat))
  return(cbind(t(dist_mat), max.col(-t(dist_mat))))#last column is their belonging cluster
}
#example dist_mat2mat(digits,matrix(c(runif(3*784, min=0, max=255)),nrow = 3, ncol = 784))
#pure distance dunction between mat 2 mat
distance_mat2mat = function(input_mat, ref_mat){
  dist_mat = apply(input_mat, 1, function(r) dist_vec2mat(r, ref_mat))
  return(t(dist_mat))
}
get_cluster = function(matrix){
  dist = apply(matrix,1,function(r) c(t(r[r[length(r)]]),r[length(r)]))
  return(t(dist))
}
#k_means function
my_kmeans = function(matrix_digits, k, N){
  min_old = Inf
  loss_record = list()#record N times loss function value
  loss_min = list()#N times minimum loss value
  for(i in 1:N){
    set.seed(i)
    mean_new = matrix(c(runif(k*784, min=0, max=255)),nrow = k, ncol = 784) #generate k cluster mean, every row is one mean for future cluster
    mean_dif = 100000000000
    loss_for_N = list()
    loss_b = list()#loss for question b
    while(mean_dif>0.001){#threshold
      mean_old = mean_new
      mean_new = matrix(c(rep(0)), nrow = k, ncol = 784)
      cluster = dist_mat2mat(digits, mean_old)
      for(j in 1:k){
        #value_cluster = cluster[,cluster[2,] == j]
        if(j %in% cluster[,k+1]){
          po_of_cluster = digits[cluster[,k+1]==j,]
          mean_of_cluster = rowMeans(t(po_of_cluster))
          mean_new[j,] = mean_of_cluster
        }
        else{
          mean_new[j,] = matrix(c(rep(1,784)),nrow=1,ncol = 784)#mean_new[j,] = mean_old[j,]
        }
      }
      mean_dif = max(matrix(rowSums(abs(mean_old-mean_new)),nrow=k,ncol=1))#the max difference between new theta and old theta
      assig_cluster = get_cluster(dist_mat2mat(digits,mean_new)) #new theta cluster assignment
      loss = sum(assig_cluster[,1])#loss function for each iteration
      loss_b = append(loss_b,loss)
    }
    loss_min = append(loss_min, loss_b[length(loss_b)][1])#N minimum loss value
    if((loss_b[length(loss_b)][[1]]) < min_old){
      final_best_cluster_par = mean_new
      final_best_cluster_assign = assig_cluster
      best_sequence_loss = loss_b
      min_old = loss_b[length(loss_b)][[1]]
    }
  }
  return(list('(a) answer parameter:' = final_best_cluster_par, '(a) answer assignment:' =final_best_cluster_assign,'(b) best sequence loss:' = best_sequence_loss,'(c) N-terminal loss:' =loss_min))
}
my_kmeans(digits,6,10)
```

#3, stopping criteria 
```{r}
#I compare the difference between the old theta and the new theta, compare by row to row, for example, 
#I compare the first row of old theta and first row of new theta, I compare the second row of old theta 
#and second row of new theta, till k row of old theta and k row of new theta, then use rowsum to get total 
#difference between old theta and new theta, I extract the maximum value of the difference vector and when
#the maximum value is bellow 0.01, I stop the while loop, since the maximum change of the point is under 0.01.
```

#4 and 5, plot
all plot for k=5
```{r}
result_5 = my_kmeans(digits, 5, 25)
cluster_5_mean = unlist(result_5['(a) answer parameter:'],use.names = FALSE)
loss_5_evolution = unlist(result_5['(b) best sequence loss:'],use.names = FALSE)
```

```{r}
# plot
term_loss = result_5['(c) N-terminal loss:']
cluster_5_mean = matrix(cluster_5_mean, nrow=5, ncol = 784)
for(i in 1:5){
  show_digit(cluster_5_mean[i,])
}
library(ggplot2)

#loss diagram
x_loss_5_evolution = c(1:length(loss_5_evolution))
ggplot(data = NULL, aes(x = x_loss_5_evolution, y = loss_5_evolution))+
  geom_point(color = 'red')+
  ylab('loss function value')+xlab('running times in while loop')+
  ggtitle('loss function evolution for k=5')

#diagram problem 5
term_loss = as.data.frame(unlist(term_loss))
ggplot(data = term_loss, aes(x = unlist(term_loss)))+
  geom_density(aes(y=..scaled..),stat = 'density',position = 'identity', alpha = 0.3)+
  xlab('terminal loss function')+
  ylab('Density')+
  ggtitle('distribution of terminal loss function values')
```

plot for k=10
```{r}
result_10 = my_kmeans(digits, 10, 25)
cluster_10_mean = unlist(result_10['(a) answer parameter:'],use.names = FALSE)
loss_10_evolution = unlist(result_10['(b) best sequence loss:'],use.names = FALSE)
```

```{r}
# plot
term_loss_10 = result_10['(c) N-terminal loss:']
cluster_10_mean = matrix(cluster_10_mean, nrow=10, ncol = 784)
for(i in 1:10){
  show_digit(cluster_10_mean[i,])
}
library(ggplot2)

#loss diagram
x_loss_10_evolution = c(1:length(loss_10_evolution))
ggplot(data = NULL, aes(x = x_loss_10_evolution, y = loss_10_evolution))+
  geom_point(color = 'red')+
  ylab('loss function value')+xlab('running times in while loop')+
  ggtitle('loss function evolution for k=10')

#diagram problem 5
term_loss_10 = as.data.frame(unlist(term_loss_10))
ggplot(data = term_loss_10, aes(x = unlist(term_loss_10)))+
  geom_density(aes(y=..scaled..),stat = 'density',position = 'identity', alpha = 0.3)+
  xlab('terminal loss function')+
  ylab('Density')+
  ggtitle('distribution of terminal loss function values when k=10')
```

plot for k=20
```{r}
result_20 = my_kmeans(digits, 20, 25)
cluster_20_mean = unlist(result_20['(a) answer parameter:'],use.names = FALSE)
loss_20_evolution = unlist(result_20['(b) best sequence loss:'],use.names = FALSE)
```

```{r}
# plot
term_loss_20 = result_20['(c) N-terminal loss:']
cluster_20_mean = matrix(cluster_20_mean, nrow=20, ncol = 784)
for(i in 1:20){
  show_digit(cluster_20_mean[i,])
}
library(ggplot2)

#loss diagram
x_loss_20_evolution = c(1:length(loss_20_evolution))
ggplot(data = NULL, aes(x = x_loss_20_evolution, y = loss_20_evolution))+
  geom_point(color = 'red')+
  ylab('loss function value')+xlab('running times in while loop')+
  ggtitle('loss function evolution for k=20')

#diagram problem 5
term_loss_20 = as.data.frame(unlist(term_loss_20))
ggplot(data = term_loss_20, aes(x = unlist(term_loss_20)))+
  geom_density(aes(y=..scaled..),stat = 'density',position = 'identity', alpha = 0.3)+
  xlab('terminal loss function')+
  ylab('Density')+
  ggtitle('distribution of terminal loss function values when k=20')
```

#6, how to choose K
I use bootstrap method to choose my k value, I assign k value with a small number like 2 or 3 and I calculate the minimum loss function value under this k value. Then I assign a new value k+1 to k, and calculate the minimum loss function value of new k value. I will continue doing this when the loss function value under new k value is lees than 95% of old k value. Once new k value's loss function is greater than 95% of the old k value's, the old k value will be my prefer number of clusters K, since the greater the k is, the more complex the computation is.

##Bonus
#7
```{r}
#function computing distance between a vector and each row of a matrix
dist_vec2mat = function(x,mat){
  apply(mat,1,function(r) sum((r - x)^2))
}



#function computing every row of input_matrix
dist_mat2mat = function(input_mat, ref_mat) {
  dist_mat = apply(input_mat, 1, function(r) dist_vec2mat(r, ref_mat))
  return(cbind(t(dist_mat), max.col(-t(dist_mat))))#last column is their belonging cluster
}
#example dist_mat2mat(digits,matrix(c(runif(3*784, min=0, max=255)),nrow = 3, ncol = 784))



#pure distance dunction between mat 2 mat
distance_mat2mat = function(input_mat, ref_mat){
  dist_mat = apply(input_mat, 1, function(r) dist_vec2mat(r, ref_mat))
  return(t(dist_mat))
}

#get prototype
get_prototype = function(cluster_matrix){
  mean = matrix(c(rep(0)),nrow = 1, ncol = 784)
  distance_number_old = Inf
  for(i in 1:dim(cluster_matrix)[[1]]){
    distance_po_vector = dist_vec2mat(cluster_matrix[i,],cluster_matrix)
    distance_number_new = sum(distance_po_vector)
    if(distance_number_new < distance_number_old){
      mean = cluster_matrix[i,]
      distance_number_old = distance_number_new
    }
  }
  return(mean)
}




get_cluster = function(matrix){
  dist = apply(matrix,1,function(r) c(t(r[r[length(r)]]),r[length(r)]))
  return(t(dist))
}
#k_medoids function
my_medoids = function(matrix_digits, k, N){
  min_old = Inf
  loss_record = list()#record N times loss function value
  loss_min = list()#N times minimum loss value
  for(i in 1:N){
    set.seed(i)
    mean_new = matrix_digits[3*(1:k),] #generate k cluster mean, extract from digits_new
    mean_dif = 100000000000
    loss_for_N = list()
    loss_b = list()#loss for question b
    while(mean_dif>0.001){#threshold
      mean_old = mean_new
      mean_new = matrix(c(rep(0)), nrow = k, ncol = 784)
      cluster = dist_mat2mat(matrix_digits, mean_old)
      for(j in 1:k){
        #value_cluster = cluster[,cluster[2,] == j]
        if(j %in% cluster[,k+1]){
          po_of_cluster = matrix_digits[cluster[,k+1]==j,]
          mean_of_cluster = get_prototype(po_of_cluster)
          mean_new[j,] = mean_of_cluster
        }
        else{
          mean_new[j,] = matrix_digits[j+10,]#mean_new[j,] = mean_old[j,]
        }
      }
      mean_dif = max(matrix(rowSums(abs(mean_old-mean_new)),nrow=k,ncol=1))#the max difference between new theta and old theta
      assig_cluster = get_cluster(dist_mat2mat(matrix_digits,mean_new)) #new theta cluster assignment
      loss = sum(assig_cluster[,1])#loss function for each iteration
      loss_b = append(loss_b,loss)
    }
    loss_min = append(loss_min, loss_b[length(loss_b)][1])#N minimum loss value
    if((loss_b[length(loss_b)][[1]]) < min_old){
      final_best_cluster_par = mean_new
      final_best_cluster_assign = assig_cluster
      best_sequence_loss = loss_b
      min_old = loss_b[length(loss_b)][[1]]
    }
  }
  return(list('(a) answer parameter:' = final_best_cluster_par, '(a) answer assignment:' =final_best_cluster_assign,'(b) best sequence loss:' = best_sequence_loss,'(c) N-terminal loss:' =loss_min))
}
# my_medoids(digits_new,5,2) example if you want to try
```

#8
k=5
```{r}
# I use the first 200 images
# mean_new is the cluster prototype
#I do not want too much output, so I just output final cluster parameters and turn it into image type, 
#you can run the code to get the full output. include all the request in question 2
#first 100 digits
digits_new = digits[1:200,]

output_5 = my_medoids(digits_new,5,25)['(a) answer parameter:']
output_5 = unlist(output_5, use.names = FALSE)
output_5 = matrix(output_5, nrow = 5 ,ncol = 784)
for(i in 1:5){
  show_digit(output_5[i,])
}
```

k = 10
```{r}
output_10 = my_medoids(digits_new,10,25)['(a) answer parameter:']
output_10 = unlist(output_10, use.names = FALSE)
output_10 = matrix(output_10, nrow = 10 ,ncol = 784)
for(i in 1:10){
  show_digit(output_10[i,])
}
```

k = 20
```{r}
#for k = 20, I try more digitsm cause the cluster number is too big
digits_new_new = digits[1:500,]
output_20 = my_medoids(digits_new_new,20,25)['(a) answer parameter:']
output_20 = unlist(output_20, use.names = FALSE)
output_20 = matrix(output_20, nrow = 20 ,ncol = 784)
for(i in 1:20){
  show_digit(output_20[i,])
}
```



